Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-06-05 performance_comparison.tex
[MGa] first pass till the end

\subsection{Robustified constraints comparison}
%It is of interest to analyze how a robustified constraint---or a combination of the two described in Sections~\ref{sec:FLC} and~\ref{sec:TLC}---influences the driving style.
We now investigate how the inclusion of different kinds of robustified constraints---individually or in combination, as described in Sections~\ref{sec:FLC} and~\ref{sec:TLC}---influences the driving style.
We compare the nominal feed-forward optimal trajectory, obtained without robustified constraints, with those resulting from a robustified track limit constraint (denoted as TLC), a robustified friction limit constraint (denoted as FLC), and a scenario where both constraints are robustified (denoted as TLC+FLC). All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.

\begin{figure}
	\centering
	\includegraphics{Fig/ol_telemetries.pdf}
	\caption{Evolution of the longitudinal speed $\De u$, wheel steering angle $\De \de$, longitudinal force $\De F_x$, and lateral deviation $\De e$ are reported for three robustification scenarios: TLC (Track Limit Constraints), FLC (Friction Limit Constraint), and TLC+FLC (both constraints enforced concurrently). Each profile is shown relative to the baseline (nominal) trajectory to highlight the effect of each constraint-handling strategy. The dashed lines in the last three panels indicate the sign of the reference signal, allowing one to determine whether a positive variation with respect to the baseline corresponds to an increase or a decrease in the absolute value. This distinction is particularly important for quantities that depend on the direction of the turn, such as the wheel steering angle $\de$ and the lateral deviation $e$, while in the case of $F_x$, it indicates whether the vehicle is accelerating or braking.}
	\label{fig:ol_telemetries}
\end{figure}

@@

The comparison between the nominal case and the three robustified cases is illustrated in Figure~\ref{fig:ol_telemetries}, in terms of the deviation of selected signals from to their nominal counterparts.
The panels display variations on the longitudinal speed $\De u$ (first panel), wheel steering angle $\De \de$ (second panel), total longitudinal force $\De F_x$ (third panel), and lateral deviation of the center of mass (CoM) from the track centerline $\De e$ (fourth panel).
Except for the $\De u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal.
These indicators are shown at three distinct vertical levels---upper, central, and lower---corresponding respectively to positive, near-zero, and negative values of the nominal signal.

@@

This shift in trajectory and longitudinal speed becomes especially evident when approaching the sharp turn at $\al \in [0.707, 0.712]$. A noticeable increase in longitudinal speed is observed around $\alpha = 0.71$ for the TLC configuration  in the $\De u$ panel. This peak is followed by a sharp drop in speed, indicating intense braking.
The TLC+FLC configuration exhibits a qualitatively similar trend in $\Delta u$, although the peak is significantly lower and the subsequent speed reduction is clearly stronger. This behavior is consistent with the variation of the total longitudinal force $\Delta F_x$ shown in the third panel.
In particular, the TLC and TLC+FLC configurations exhibit a higher longitudinal force while approaching the sharp turn, in the curvilinear abscissa interval $\al\in\left[0.707, 0.712\right]$, resulting in a higher longitudinal speed.
Subsequently, these two configurations apply a lower longitudinal force during braking compared to the nominal case, indicating more intense braking. This behavior explains the sharp reduction in longitudinal speed observed thereafter.

@@

Later in the sector, a further deviation in driving style appears during the high-speed turn, specifically around $\alpha \in [0.75, 0.76]$.
As shown in the $\Delta F_x$ panel, the combined configuration TLC+FLC needs to reduce the longitudinal force within this section to meet both the constraints.
This adaptation is necessary because the TLC forces the vehicle to follow a wider trajectory, which entails greater lateral acceleration and, consequently, higher lateral force demand. To satisfy the FLC under the resulting increase in tire utilization, the available accelerating force must be reduced.
Based on the nominal force sign indicator, this reduction is achieved by partially lifting the throttle.

%---Axle saturation
After analyzing the variations of these four key signals with respect to the nominal solution, we now turn our attention to the evaluation of axle saturation, which provides insight into how close each configuration operates to the tire grip limits.
In accordance with the FLC formulation introduced in~\eqref{eq:adherence_constraint}, we define the \emph{axle saturation ratio} $S_j$ as
\begin{equation}
	S_j = \frac{ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2}{Z_j^2(\bx,\bu)}.
\end{equation}
This quantity indicates how close each configuration operates to the friction limit; $S_j = 1$ when the point $\left(X_j, Y_j\right)$ lies exactly on the friction ellipse, i.e., under full saturation.
Figure~\ref{fig:ol_saturation} shows the values of $S_j$ for the front axle ($j=1$, first panel) and the rear axle ($j=2$, rear panel) for the previously introduced configurations: the nominal case (black dashed line) and the three robustified cases (color-coded as before).

\begin{figure}
	\centering
	\includegraphics{Fig/ol_saturation.pdf}
	\caption{Tire usage for front axle (first panel) and rear axle (second panel) over curvilinear abscissa $\al$. The plotted lines represent the quantity $\left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}$, where $j=1$ denotes the front axle and $j=2$ the rear axle. The dashed black line corresponds to the reference (nominal) configuration, while the green, red, and blue lines represent the robust configurations TLC, FLC, and TLC+FLC, respectively.}
	\label{fig:ol_saturation}
\end{figure}

First, we observe that the FLC and TLC+FLC configurations (red and blue lines) never reach full saturation, i.e., $S_j=1$, due to the presence of the back-off term, which enforces a safety margin from the friction limit.
Second, we point out how the different trajectories followed by the TLC and TLC+FLC configurations (green and blue lines), compared to the nominal and FLC cases, entail distinct ground force demands, which are clearly visible here in terms of the $S_j$ ratio.
This behavior emerges around the sharp turn and becomes even more evident in the high-speed turn.
Specifically, in the interval $\alpha \in [0.745, 0.762]$, the TLC and TLC+FLC configurations begin to experience high axle loads approximately 12\,m before the nominal trajectory and sustain them for an additional 4.6\,m (TLC, green line) and 9.2\,m (TLC+FLC, blue line) beyond the nominal reference.

@@

The left panel of Figure~\ref{fig:traj_strings} shows random evolutions (red lines) for an LQR-controlled vehicle stabilized to follow the nominal optimal trajectory (black line). No robust constraints are introduced, and the controller implements the feedback policy from step 2 of Subsection~\ref{sec:LQR}. The right panel, instead, shows the random evolutions (red lines) for the robustified LQR-based controller obtained in the step 3 of the closed-loop approach described in Subsection~\ref{sec:CLrobustified}. The corresponding robustified mean trajectory is shown in black.
The comparison between the panels shows that the robustified controller is significantly more effective at satisfying the TLC, resulting in safer and more reliable trajectory tracking.

@@

\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_strings.pdf}
	\caption{Simulations with random initial conditions and Gaussian noise realizations. In the left panel, the reference trajectory is the non-robust one, tracked using an LQR controller designed around it. In the right panel, the reference is a robust trajectory obtained using the closed-loop optimization method, and it is tracked using the corresponding optimized controller produced by the same method.}
	\label{fig:traj_strings}
\end{figure}

>>>>

\subsection{Robustified constraints comparison}
%It is of interest to analyze how a robustified constraint---or a combination of the two described in Sections~\ref{sec:FLC} and~\ref{sec:TLC}---influences the driving style.
We now investigate how the inclusion of different kinds of robustified constraints --- individually or in combination, as described in Sections~\ref{sec:FLC} and~\ref{sec:TLC} --- influences the driving style.
We compare the nominal feed-forward optimal trajectory, obtained without robustified constraints, with those resulting from a robustified track limit constraint (denoted as TLC), a robustified friction limit constraint (denoted as FLC), and a scenario where both constraints are robustified (denoted as TLC+FLC). All the optimization are obtained with the open-loop method described in Section~\ref{sec:open_loop_planning}, setting $H=4$, and $\ga^\textrm{TLC}=\ga^\textrm{FLC}=1.28$, corresponding to a 90\% probability of satisfying both constraints.

\begin{figure}
	\centering
	\includegraphics{Fig/ol_telemetries.pdf}
\caption{Variations of longitudinal speed $\Delta u$, steering angle $\Delta\delta$, longitudinal force $\Delta F_x$, and lateral deviation $\Delta e$ relative to the nominal trajectory for three schemes --- TLC (track-limit), FLC (friction-limit), and TLC+FLC (both).
Dashed lines in the last three panels mark the sign of the reference signal, indicating whether a positive variation increases or decreases the absolute value (crucial for the turn-dependent variables $\delta$ and $e$ and for distinguishing traction from braking in $F_x$).}

%\caption{Evolution of the longitudinal speed $\De u$, wheel steering angle $\De \de$, longitudinal force $\De F_x$, and lateral deviation $\De e$ are reported for three robustification scenarios: TLC (Track Limit Constraints), FLC (Friction Limit Constraint), and TLC+FLC (both constraints enforced concurrently). Each profile is shown relative to the baseline (nominal) trajectory to highlight the effect of each constraint-handling strategy. The dashed lines in the last three panels indicate the sign of the reference signal, allowing one to determine whether a positive variation with respect to the baseline corresponds to an increase or a decrease in the absolute value. This distinction is particularly important for quantities that depend on the direction of the turn, such as the wheel steering angle $\de$ and the lateral deviation $e$, while in the case of $F_x$, it indicates whether the vehicle is accelerating or braking.}
	\label{fig:ol_telemetries}
\end{figure}

@@

The comparison between the nominal case and the three robustified cases is illustrated in Figure~\ref{fig:ol_telemetries}, in terms of the deviation of selected signals from to their nominal counterparts.
The panels display variations on the longitudinal speed $\De u$ (first panel), wheel steering angle $\De \de$ (second panel), total longitudinal force $\De F_x$ (third panel), and lateral deviation of the center of mass (CoM) from the track centerline $\De e$ (fourth panel).
Except for the $\De u$ panel, each plot includes dashed black lines that indicate the sign of the nominal signal.
These indicators are shown at three distinct vertical levels --- upper, central, and lower --- corresponding respectively to positive, near-zero, and negative values of the nominal signal.

@@

This shift in trajectory and longitudinal speed becomes especially evident when approaching the sharp turn for $\al \in [0.707, 0.712]$. A noticeable increase in longitudinal speed is observed around $\alpha = 0.71$ for the TLC configuration  in the $\De u$ panel. This peak is followed by a sharp drop in speed, indicating intense braking.
The TLC+FLC configuration exhibits a qualitatively similar trend in $\Delta u$, although the peak is significantly lower and the subsequent speed reduction is clearly stronger. This behavior is consistent with the variation of the total longitudinal force $\Delta F_x$ shown in the third panel.
\textcolor{red}{In particular, the TLC and TLC+FLC configurations exhibit a higher longitudinal force while approaching the sharp turn, in the curvilinear abscissa interval $\al\in\left[0.707, 0.712\right]$, resulting in a higher longitudinal speed.
Subsequently, these two configurations apply a lower longitudinal force during braking compared to the nominal case, indicating more intense braking. This behavior explains the sharp reduction in longitudinal speed observed thereafter.}

@@

Later in the sector, a further deviation in driving style appears during the high-speed turn, specifically around $\alpha \in [0.75, 0.76]$.
As shown in the third panel ($\Delta F_x$), the combined configuration TLC+FLC needs to reduce the longitudinal force within this sector to meet both the constraints.
This adaptation is necessary because the TLC forces the vehicle to follow a wider trajectory, which entails greater lateral acceleration and, consequently, higher lateral force demand. To satisfy the FLC under the resulting increase in tire utilization, the available accelerating force must be reduced.
Based on the nominal force sign indicator, this reduction is achieved by partially lifting the throttle.

%---Axle saturation
After analyzing the variations of these four key signals with respect to the nominal solution, we now turn our attention to the evaluation of axle saturation $S_j(\bx,\bu)$ defined in~\eqref{eq:axle_saturation}, which provides insight into how close each configuration operates to the tire grip limits.

Figure~\ref{fig:ol_saturation} shows the values of $S_j(\bx,\bu)$ for the front axle ($j=1$, first panel) and the rear axle ($j=2$, second panel) for the previously introduced configurations: the nominal case (black dashed line) and the three robustified cases TLC, FLC and TLC+FLC (color-coded as before).

\begin{figure}
	\centering
	\includegraphics{Fig/ol_saturation.pdf}
	\caption{Axle saturation $S_j(\bx,\bu)$ for front axle (j=1, first panel) and rear axle (j=2, second panel) with respect to the curvilinear abscissa $\al$. The dashed black line corresponds to the reference (nominal) configuration, while the green, red, and blue lines represent the robustified configurations TLC, FLC, and TLC+FLC, respectively.}
	\label{fig:ol_saturation}
\end{figure}

Firstly, we observe that the FLC and TLC+FLC configurations (red and blue lines) never reach full saturation ($S_j=1)$ due to the presence of the back-off term, which enforces a safety margin from the friction limit.
Secondly, we point out how the different trajectories followed by the TLC and TLC+FLC configurations (green and blue lines), compared to the nominal and FLC cases, entail distinct ground force demands, which are clearly visible here in terms of the $S_j$ ratio.
This behavior emerges around the sharp turn and becomes even more evident in the high-speed turn.
Specifically, in the interval $\alpha \in [0.745, 0.762]$, the TLC and TLC+FLC configurations begin to experience high axle loads approximately 12\,m before the nominal trajectory and sustain them for an additional 4.6\,m (TLC, green line) and 9.2\,m (TLC+FLC, blue line) beyond the nominal reference.

@@

The left panel of Figure~\ref{fig:traj_strings} shows random evolutions (red lines) for an LQR-controlled vehicle stabilized to follow the nominal optimal trajectory (black line). No robust constraints are introduced, and the controller implements the feedback policy from Step 2 of Subsection~\ref{sec:LQR}. The right panel, instead, shows the random evolutions (red lines) for the robustified LQR-based controller obtained in the Step 3 of the closed-loop approach described in Subsection~\ref{sec:CLrobustified}. The corresponding robustified mean trajectory is shown in black.
The comparison between the panels shows that the robustified controller is significantly more effective at satisfying the TLC, resulting in safer and more reliable trajectory tracking.

@@

\begin{figure}
	\centering
	\includegraphics{Fig/olcl_traj_strings.pdf}
	\caption{Simulations with random initial conditions and Gaussian noise realizations. In the left panel, the reference trajectory (black curve) is the non-robust one, tracked using an LQR controller designed around it as per Step 2 of Subsection~\ref{sec:LQR}. In the right panel, the reference (black curve) is a robust trajectory obtained using the closed-loop optimization method as per Step 3 of Subsection~\ref{sec:CLrobustified}, which is tracked using the corresponding optimized controller.}
	\label{fig:traj_strings}
\end{figure}

### 2025-06-05 stochastic_vehicle_dynamics.tex
[MGa] first pass till the end

The base version of the constraint for each axle can be expressed as a function of $\bx$ and $\bu$, as follows:
\begin{equation}
	h^\textrm{FLC}_j(\bx,\bu) = \left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}-1\leq0\qquad j=1,2.\label{eq:adherence_constraint}
\end{equation}
where the superscript FLC denotes friction limit constraint, and the subscript $j$ indicates the axle index, with $j=1$ referring to the front axle and $j=2$ to the rear axle. Here, $X_j$ and $Y_j$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z_j$ represents the total vertical load acting on the axle.
As can be seen from~\eqref{eq:adherence_constraint}, the points $\left(X_j,Y_j\right)$ are constrained to lie within an ellipse whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$. The constraint, without the back-off term, is represented in the left panel of Figure~\ref{fig:robust_constraints} by the solid line ellipse in the plane $X_j-Y_j$.

>>>>

The base version of the constraint for each axle can be expressed as the following inequality depending on $\bx$ and $\bu$
\begin{equation}
	h^\textrm{FLC}_j(\bx,\bu) =  S_j(\bx,\bu) - 1 \leq 0, \qquad{(j=1,2)}
\label{eq:adherence_with_axle_saturation}
\end{equation}
where we introduced the \emph{axle saturation ratio} $S_j(\bx,\bu)$ as follows
\begin{equation}
	S_j(\bx,\bu) = \frac{ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2}{Z_j^2(\bx,\bu)}.
\label{eq:axle_saturation}
\end{equation}
In~\eqref{eq:adherence_with_axle_saturation}, the superscript FLC denotes friction limit constraint, and $j=1,2$ refer the front and rear axle, respectively. In~\eqref{eq:axle_saturation}, $X_j$ and $Y_j$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z_j$ represents the total vertical load acting on the axle.

From~\eqref{eq:axle_saturation} it is evident that $S_j(\bx,\bu)\in [0,1]$ indicates how close each configuration operates to the friction limit: $S_j=0$ when the overall grip demand is zero, while $S_j = 1$ when the point $\left(X_j, Y_j\right)$ lies exactly on the friction ellipse, i.e., under full saturation.
The points $\left(X_j,Y_j\right)$ are constrained to lie within an ellipse whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$. The constraint, without the back-off term, is represented in the left panel of Figure~\ref{fig:robust_constraints} by the solid line ellipse in the plane $X_jY_j$.
